Board Game Recommendation

### 2015-03-26 paper.Rnw
just a few comments

\begin{table}[ht]
\centering
\begin{tabular}{|l|c|}
\hline
Category &  This covers the broad categories a game\\
& might fall into (ex.: Humor, Puzzle, Sports, etc.) \\
\hline
Family &  Some games a naturally part of a "family" \\
&  that share a name or common element \\
& (ex.: Animals:Bats, Ancient Wars Series, Hello Kitty)\\
\hline
Mechanic &  This is what mechanisms the game uses\\
& (ex.: Dice Rolling, Card Drafting, Worker Placement)\\
\hline
Subdomain & More general than Category, but loosely bins games\\
& (ex.: Family Games, Strategy Games, etc.)\\
\hline
\end{tabular}
\caption{\label{tab:categories} Overview of the  relevant characteristics of board games considered in this project.}
\end{table}

@@

To keep the information as compact as possible, individual players and games are referred to by an ID number, which corresponds to a row in one of the key tables. These key tables contain information about the players and game that we don't necessarily need to include with each rating or classifier. This greatly reduces the size of the database needed to store and work with the information.
<<player_key, echo = FALSE, results = 'asis'>>=
xtable(paper_stuff$player_key, caption = c("The Player Key relates important info about the individual to the userid", "Player Key Table"))
@

@@

There are a few things that make sense to model first. Certain games are more popular than other games, and certain people will naturally be more or less generous than others in their ratings. These two effects are very important to measure, as the first will allow us to provide estimates for people that have rated no games (and therefore we know nothing about), and the second will allow us to fine tune estimates for people who have rated many games. We will implement these in a similar manner to the way that BigChaos implemented the Global Effects model \citep{bigchaos}.
\hh{XXX the next clause needs an end to be a sentence :)}
While they implemented 14 steps including Movie Effect, User Effect, and those variables crossed with variables like time since rating, average rating for the movie, production year, standard deviations, as well as others. 
\hh{XXX explain RMSE; make sure in the next paragraph to always point out what is yours and what is netflix. You know all those distinctions, but for any other reader it is not that obvious.}
From the RMSE of their probe set, the difference between using only user/movie effcts and using all 14 effects was only .02, a decrease of about 2\%. Noting this negligible increase in accuracy and recognizing my lack of significant computing resources, we will only use user and game effects.

After we fit the general effects, we wish to model the remaining residuals, ideally in some way that gives us informative numeric summaries about the individual users in the process. In addition, we'd also like to compare a users ratings to other users, and hopefully use the behavior of simiar users to predict new ratings for someone. A tempting first step is to, for a particular game, find other people who have rated that game who have rated games that you've rated. Then, you could find people who rated games similarly to you, and use their ratings of the new game as an estimate. This often doesnt work, as sometimes the number of people who have rated the same games as you can be relatively low, if not 0 (show a graph for this part).

@@

\subsection{Model Structure}
The specifics of the model follow very closely to those outlined in the Global Effects and knn sections of the BigChaos paper [need bib]. Here, we cover the specifics of both.

Note for editing: I know this model is a mess at the moment, just getting the general ideas down on paper. 
\hh{this is not nearly as messy as you make it out to be, but you need to make sure to define every little piece of notation as you, because otherwise your reader will be lost.}
\subsubsection{Global Effects}
The general idea behind global effects is to account for the natural biases present in particular games or players. For example, we might have two players who both play and enjoy the same game the same amount. However, one person might give this game a score of 10 while the other gives it a score of 9. Despite their overall enjoyment being the same, perhaps the latter player just tends to be a harsher rater in general. The global effect for this person would then be lower than that of the much more cheery fellow who gave the game a 10. Similarly, this effect can be present for games due to any number of reasons.

@@

\hh{XXX: generally:  no future tense in the write up. The CC is in the here and now :)
}

\hh{XXX it is $\theta_u$, not $\theta$. }
\hh{XXX Why do you call $\alpha$ a fitting parameter and not a shrinking parameter?}

@@

\subsubsection{Classifier Aggregation}
After finding the global effects for the games and the users, we're left with residuals for every rating that every user has made. These residuals represent the last of what differentiate the users based on their preferences for the nuances of each game. We need a way to quantify these nuances, and then attempt to model these residual so we can get a better understanding of each individual user. Each game in the dataset can be described by things like what mechanics it has, or what family of game it falls under, when it was published, etc. All in all, there are up to 1,851 different quantifiable ways that were chosed to descibe games. For example, the game Dragonmaster was published in 1981, has a playingtime of 30 minutes, is under the family "Animals: Dragons", falls under the categories "Card Game", "Fantasy Game", is in the subdomain "Strategy Games", uses the mechanic "Trick-taking", and seats 3 or 4 players. Each one of these will be treated as a classifier for the game.

In order to model the remaining residuals, we'll give each player a score based on how they rated games with similar classifiers; that is, how did they rate games with the "Trick-taking" mechanic, or how did they rate "Card Games"?

@@

where the sum is over all residuals that belong to a game that can be described by classifier $z$. For example, if the classifier was the mechanic "Trick-taking" and the user was $u = 3$, the classifier score would be the mean of all of the $\epsilon_{3,i}^{(3)}$ for \textit{all} games that the user has rated that have the "Trick-taking" mechanic. If its the case that a user has rated no games with a particular classifier, we'll leave it as an unobserved value.

\subsubsection{k-nearest-neighbors}
Now we have a few classifier values defined for each user who has rated at least one game. In order to create predictions for other games that the user hasn't played, we'll find the correlation between players based on their classifier values. For each pair of players, the correlation between them can be found by finding all of the classifiers that have scores for both players, and then finding the correlation between those values. To make sure that we don't find high correlations between players who only have very few classifier scores in common, we'll on consider a correlation between players if they have more than 6 common classifier scores.

@@

\subsubsection{Optimizing $\alpha$}
To optimize the $\alpha$ values, we'll follow the model fitting procedure up to the point in which the paramter is used, estimate the model parameter using a grid of possible values for the turning parameter, and then find which value for the tuning parameter yields the model parameter estimate which has the lowest root mean square error (RMSE). RMSE is calculated by finding the difference between the rating and the rating that is estimated using only the global effects parameter(s).

The first $\alpha$ to be tuned is $\alpha_{game}$. The data is randomly divided into 10 separate sections of roughly equal size. Then for each of the 10 sections, we'll treat the section as a test set and the other 9 to use as a training set. $\widehat{\theta}_g$ is fit first, and then from a grid of $\alpha_{game}$ values (where $\alpha_{game} \in [1,10000]$), a value of $\widehat{\theta}_i$ is produced for each of the $\alpha_{game}$ in the grid. For each of the $\widehat{\theta}_i$ that are estimated, we'll use the section that wasn't in the training set, the RMSE is calculated by:
\begin{equation}
\text{RMSE} = \frac{ \sum r_{u,i}^{(1)} - (\theta_g + \theta_i)}{n}
\end{equation}

where we sum over all rated games in the test set, and $n$ is the number of ratings in the test set. The value of $\alpha_{game}$ that has the lowest RMSE will be used. in figure ~\ref{fig:gamealpha}, we see the optimal value for $\alpha_{game}$ sits in a nice trough, with the minimum at $\alpha_{game} = 1.802303$.

\begin{center}
\begin{figure}
<<game_alpha_optim, echo = F, fig.width = 4, fig.height = 4 >>=
load("game_graph")
game_graph
@
\caption{\label{fig:gamealpha} RMSE of candidate $\alpha_{game}$ values.}
\end{figure}
\end{center}

@@

Once again, the value of $\alpha_{player}$ with the lowest RMSE will be the value used, which as can be seen in ~\ref{fig:playeralpha} is $\alpha_{player} = 2.022021$.
\begin{figure}
<<player_alpha_optim, echo = F, fig.width = 4, fig.height = 4>>=
load("player_graph")
player_graph
@
\caption{\label{fig:playeralpha} RMSE of candidate $\alpha_{player}$ values.}
\end{figure}

@@

\begin{figure}
<<hists, echo = F, fig.width = 4, fig.height = 4>>=
load("game_n_rats")
load("plyr_n_rats")
game_n_plot
plyr_n_plot
@
\caption{\label{fig:ratn} Frequency of number of ratings for players and games.}
\end{figure}

@@

% don't box citations, add space between multiple citations, separate with ;
\def\@citex[#1]#2{\if@filesw\immediate\write\@auxout{\string\citation{#2}}\fi
  \def\@citea{}\@cite{\@for\@citeb:=#2\do
    {\@citea\def\@citea{; }\@ifundefined
       {b@\@citeb}{{\bf ?}\@warning
       {Citation `\@citeb' on page \thepage \space undefined}}%
{\csname b@\@citeb\endcsname}}}{#1}}

>>>>

\begin{table}[ht]
\centering
\begin{tabular}{|l|c|}
\hline
Category &  This covers the broad categories a game\\
& might fall into (ex.: Humor, Puzzle, Sports, etc.) \\
\hline
Family &  Some games a naturally part of a ``family" \\
&  that share a name or common element \\
& (ex.: Animals:Bats, Ancient Wars Series, Hello Kitty)\\
\hline
Mechanic &  This is what mechanisms the game uses\\
& (ex.: Dice Rolling, Card Drafting, Worker Placement)\\
\hline
Subdomain & More general than Category, but loosely bins games\\
& (ex.: Family Games, Strategy Games, etc.)\\
\hline
\end{tabular}
\caption{\label{tab:categories} Overview of the  relevant characteristics of board games considered in this project.}
\end{table}

@@

To keep the information as compact as possible, individual players and games are referred to by an ID number, which corresponds to a row in one of the key tables. These key tables contain information about the players and game that we do not necessarily need to include with each rating or classifier. This greatly reduces the size of the database needed to store and work with the information.
<<player_key, echo = FALSE, results = 'asis'>>=
xtable(paper_stuff$player_key, caption = c("The Player Key relates important info about the individual to the userid", "Player Key Table"))
@

@@

There are a few things that make sense to model first. Certain games are more popular than other games, and certain people will naturally be more or less generous than others in their ratings. These two effects are very important to measure, as the first will allow us to provide estimates for people that have rated no games (and therefore we know nothing about), and the second will allow us to fine tune estimates for people who have rated many games. We will implement these in a similar manner to the way that BigChaos implemented the Global Effects model \citep{bigchaos}.
\hh{XXX the next clause needs an end to be a sentence :)}
While they implemented 14 steps including Movie Effect, User Effect, and those variables crossed with variables like time since rating, average rating for the movie, production year, standard deviations, as well as others. 
\hh{XXX explain RMSE; make sure in the next paragraph to always point out what is yours and what is netflix. You know all those distinctions, but for any other reader it is not that obvious.}
From the RMSE of their probe set, the difference between using only user/movie effcts and using all 14 effects was only .02, a decrease of about 2\%. Noting this negligible increase in accuracy and recognizing my lack of significant computing resources, we will only use user and game effects. \hh{XXX What is your computing power? How much further could you go with (how much?) more? -- How much more accuracy would you hope for with more power? }

After we fit the general effects, we wish to model the remaining residuals, ideally in some way that gives us informative numeric summaries about the individual users in the process. In addition, we'd also like to compare a users ratings to other users, and hopefully use the behavior of simiar users to predict new ratings for someone. A tempting first step is to, for a particular game, find other people who have rated that game who have rated games that you've rated. Then, you could find people who rated games similarly to you, and use their ratings of the new game as an estimate. This often does not work, as sometimes the number of people who have rated the same games as you can be relatively low, if not 0 (show a graph for this part).

@@

\subsection{Model \hh{Sequence}}
The specifics of the model follow very closely to those outlined in the Global Effects and knn sections of the BigChaos paper [need bib]. Here, we cover the specifics of both.

Note for editing: I know this model is a mess at the moment, just getting the general ideas down on paper.

\hh{XXX Describe the sequential nature of the models in the next subsections here.}

\subsubsection{Global Effects}
The general idea behind global effects is to account for the natural biases present in particular games or players. For example, we might have two players who both play and enjoy the same game the same amount. However, one person might give this game a score of 10 while the other gives it a score of 9. Despite their overall enjoyment being the same, perhaps the latter player just tends to be a harsher rater in general. The global effect for this person would then be lower than that of the much more cheery fellow who gave the game a 10. Similarly, this effect can be present for games due to any number of reasons.

@@

@@

\subsubsection{Classifier Aggregation}
After finding the global effects for the games and the users, we're left with residuals for every rating that every user has made. These residuals represent the last of what differentiate the users based on their preferences for the nuances of each game. We need a way to quantify these nuances, and then attempt to model these residual so we can get a better understanding of each individual user. Each game in the dataset can be described by things like what mechanics it has, or what family of game it falls under, when it was published, etc. All in all, there are up to 1,851 different quantifiable ways that were chosed to descibe games. For example, the game Dragonmaster was published in 1981, has a playingtime of 30 minutes, is under the family ``Animals: Dragons", falls under the categories ``Card Game", ``Fantasy Game", is in the subdomain ``Strategy Games", uses the mechanic ``Trick-taking", and seats 3 or 4 players. Each one of these will be treated as a classifier for the game.

In order to model the remaining residuals, we'll give each player a score based on how they rated games with similar classifiers; that is, how did they rate games with the ``Trick-taking" mechanic, or how did they rate ``Card Games"?

@@

where the sum is over all residuals that belong to a game that can be described by classifier $z$. For example, if the classifier was the mechanic ``Trick-taking" and the user was $u = 3$, the classifier score would be the mean of all of the $\epsilon_{3,i}^{(3)}$ for \textit{all} games that the user has rated that have the ``Trick-taking" mechanic. If its the case that a user has rated no games with a particular classifier, we'll leave it as an unobserved value.

\subsubsection{k-nearest-neighbors}
Now we have a few classifier values defined for each user who has rated at least one game. In order to create predictions for other games that the user hasn't played, we'll find the correlation between players based on their classifier values. For each pair of players, the correlation between them can be found by finding all of the classifiers that have scores for both players, and then finding the correlation between those values. To make sure that we do not find high correlations between players who only have very few classifier scores in common, we'll on consider a correlation between players if they have more than 6 common classifier scores.

@@

\subsubsection{Optimizing tuning parameters $\alpha$}
\hh{XXX avoid using symbols without also calling them by their names ... i.e. say tuning parameter $\alpha$ rather than just $\alpha$.}

To optimize the $\alpha$ values, we follow the model fitting procedure up to the point in which the paramter is used, estimate the model parameter using a grid of possible values for the turning parameter, and then find which value for the tuning parameter yields the model parameter estimate which has the lowest root mean square error (RMSE).
\hh{XXX back this definition of RMSE up with a formula and move it to the first time where you mention RMSE. }
RMSE is calculated by finding the difference between the rating and the rating that is estimated using only the global effects parameter(s).

\hh{What does 'the first'meane here? Does it mean that there is a given order, which we have to follow, or did you decide to first tune the game parameter? Does it matter which parameter is tuned first?}
The first $\alpha$ to be tuned is $\alpha_{game}$.
\hh{XXX What you are describing here is called ten-fold cross validation. }
The data is randomly divided into 10 separate sections of roughly equal size. Then for each of the 10 sections, we'll treat the section as a test set and the other 9 to use as a training set. $\widehat{\theta}_g$ is fit first, and then from a grid of $\alpha_{game}$ values (where $\alpha_{game} \in [1,10000]$), a value of $\widehat{\theta}_i$ is produced for each of the $\alpha_{game}$ in the grid. For each of the $\widehat{\theta}_i$ that are estimated, we'll use the section that wasn't in the training set, the RMSE is calculated by:
\begin{equation}
\text{RMSE} = \frac{ \sum r_{u,i}^{(1)} - (\theta_g + \theta_i)}{n}
\end{equation}

where we sum over all rated games in the test set, and $n$ is the number of ratings in the test set. The value of $\alpha_{game}$ that has the lowest RMSE will be used. in figure ~\ref{fig:gamealpha}, we see the optimal value for $\alpha_{game}$ sits in a nice trough, with the minimum at $\alpha_{game} = 1.802303$. \hh{XXX do you really need that many significant digits?}

\begin{center}
\begin{figure}
\centering
<<game_alpha_optim, echo = F, fig.width = 4, fig.height = 4, out.width='0.5\\textwidth'>>=
load("game_graph")
game_graph
@
\caption{\label{fig:gamealpha} RMSE of candidate $\alpha_{game}$ values. \hh{XXX there's a typo in the x label }}
\end{figure}
\end{center}

@@

Once again, the value of $\alpha_{player}$ with the lowest RMSE will be the value used, which as can be seen in ~\ref{fig:playeralpha} is $\alpha_{player} = 2.022021$.
\begin{figure}
\centering
<<player_alpha_optim, echo = F, fig.width = 4, fig.height = 4, out.width='0.5\\textwidth'>>=
load("player_graph")
player_graph
@
\caption{\label{fig:playeralpha} RMSE of candidate $\alpha_{player}$ values. \hh{typo in X}}
\end{figure}

@@

\begin{figure}
\centering
<<hists, echo = F, fig.width = 4, fig.height = 4, out.width='0.45\\textwidth', fig.show='hold'>>=
load("game_n_rats")
load("plyr_n_rats")
game_n_plot
plyr_n_plot
@
\caption{\label{fig:ratn} Frequency of number of ratings for games (left) amd players (right). \hh{A log transformation on the frequencies might be more informative. Make sure to emphasize it, if you show transformed data.}}
\end{figure}

@@

% do not box citations, add space between multiple citations, separate with ;
\def\@citex[#1]#2{\if@filesw\immediate\write\@auxout{\string\citation{#2}}\fi
  \def\@citea{}\@cite{\@for\@citeb:=#2\do
    {\@citea\def\@citea{; }\@ifundefined
       {b@\@citeb}{{\bf ?}\@warning
       {Citation `\@citeb' on page \thepage \space undefined}}%
{\csname b@\@citeb\endcsname}}}{#1}}
